Kindly Bent to Free Us

### 2019-07-10 kindly/lang.tex
checkpoint-semantics

\clearpage
\subsection{Semantics}
\label{sem}

\input{internal-grammar}
\input{fig-reduction}
\input{fig-code}

>>>>

\clearpage
\subsection{Semantics}
\label{sem}
\input{internal-grammar}

\lstMakeShortInline[style=rule]@

In this section, we present the dynamics of \lang in big-step
functional
style~\cite{siek13:_type_safet_three_easy_lemmas,DBLP:conf/esop/OwensMKT16,
  DBLP:conf/popl/AminR17}. It is given as a function
@eval \Store \Perm \VEnv i e@ 
that manipulates the semantic objects defined in
\cref{fig:syntax-internal-language}. 
To demonstrate resource handling, the semantics comes with an abstract
notion of linear resources that can be created, destroyed, observed
(through unrestricted borrows), and
updated (through affine borrows).

Expressions $e$ need to be elaborted with some kind and constraint
annotations. Polymorphic functions give rise to elaborated
abstractions which abstract over kind variables $\Multi\kappa$ and
which are tagged with a kind $k$ indicating their usage restriction
(linear, affine, etc). Their use
gives rise to elaborated application where the instantiation of these
kind variables is given. Pairs also come with a kind tag $k$
indicating the usage restriction.

Addresses $\Addr$ are composed of a raw location $\Loc$, which is just
a pointer into a store, and a stack of modifiers that indicates the
borrows that have been taken from the original object. From the raw
location, we may take affine borrows and reborrowed. Once we have
taken an unrestricted borrow (from a raw location or a borrowed one),
then we can take further unrestricted borrows from it, but no more
affine ones.

A permission $\Perm$ is a set of addresses. During evaluation it
contains the set of addresses that may be accessed. For each raw
location, there is only ever one address inside of $\Perm$.

A result $r$ is either an address or a constant (e.g., a number). All
non-trivial values are boxed in the sense that they are represented by
the address of their actual content.

A value environment $\VEnv$ is a mapping from variables to results.

A storable $W$ is a particular object that can be stored. There are five
kinds of storable. A poly closure represents a polymorphic
function. It consists of an environment and the components of an
elaborated abstraction. Similarly, a closure represents a monomorphic
function and it elides kind variables and the constraint. A pair
consists of a kind and two results. A resource is just an indirection
to another result and the $\blob $ fills a released location.

Finally, the store $\Store$ is a mapping from raw locations to
storables. It comes with a function
@salloc : store -> storable -> (loc * store)@ such that
@salloc delta w@ allocates a free location in @delta@ and fills it with
storable $w$. It returns the location and the updated store.

The evaluation function is indexed by a step count so that each
invocation is guaranteed to terminate either with an error, a timeout,
or a meaningful result. Hence its return type is a monad
\lstinline/'a sem/ which combines error reporting, timeout, and a
result of type \lstinline/'a/:
\begin{lstlisting}
type 'a sem = Error of string | TimeOut | Ok of 'a
# eval : store -> perm -> venv -> int -> exp -> (store * perm * result) sem
\end{lstlisting}
It evaluates the given expression in the context of an initial store, a
permission to use addresses in the store, a value environment, and a
step count. If successful, it returns the final store, the remaining
permissions, and the actual result.

\input{fig-code}

We give some excerpts of the definition of @eval@ in
\cref{fig:big-step-interpretation} and leave the full 
definition for the appendix~\cref{sec:semant-defin}. The appendix also
contains the semantics 
in traditional notation, which does not make the monadic nature of
evaluation explicit.

The definition uses concrete OCaml syntax with some pretty
printing. The pervasive \lstinline/let*/ operator acts as monadic bind
for the \lstinline/sem/ monad. The function header of @eval@ checks
whether time is up and otherwise proceeds processing the expression.

The @Varinst@ case corresponds to the elaborated application. It
obtains the variable's value, checks that it is a location, checks the
permission (the @let*?@ clause), obtains the storable $w$ at that
location, and checks that it is an STPOLY. Then it updates the
permission: if the STPOLY closure is unrestricted, then the location
remains in the permission set, otherwise it is removed. Finally, we
allocate a new monomorphic closure, add it to the permissions, and
return the pointer as the result along with the updated store and
permissions.

The @App@ case implements function application. Compared to the syntax
it comes with an annotation @sp@ which contains evidence of the
environment splitting in the application rule. This evidence takes the
form of two sets of variables from the domain of @gamma@.

To execute an application, we first apply the splitting to @gamma@ and
evaluate subterm @e_1@ with its part of the environment and the
decremented timer @i'@. The result must be a location that we are
permitted to use. Moreover, there must be a monomorphic STCLOS stored
at that location. The permission to further use this closure is
withdrawn unless the closure is unrestricted. Finally, we evaluate the
argument, then the function body, and return its result.

The @Pair@ case does not present any new features, but it
corresponds to the \TirName{Pair} typing rule.

The @Region@ case implements a region. It obtains the address for @x@,
the suspended binding, and extends it with the intended borrow
$\BORROW$. This extension may fail if we try to take an affine borrow
of an unrestricted borrow. Next, we rebind @x@ to the borrow's
address, extend the permission accordingly, and execute the region's
body.  Finally, we withdraw the permission and return the result.

The @Borrow@ case obtains the address for @x@, checks that it is a
borrow of the correct mode $\BORROW$ and whether it is permitted to
use it. It just returns the address.

\lstDeleteShortInline@
